---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load necessary packages:
```{r}
library(dplyr)
library(ggplot2)
```

Unzip raw data file if necessary:
```{r}
if(!file.exists("activity.csv")) {
    if(!file.exists("activity.zip")) {
        print("Source data not in working directory.")
    } else unzip("activity.zip")
}
```

Load raw data and format date variable:
```{r}
activity <- tbl_df(read.csv("activity.csv"))
activity$date <- as.Date(activity$date, format="%Y-%m-%d")
```


## What is mean total number of steps taken per day?
Calculate total steps per day:
```{r}
activity_by_day <- group_by(activity, date)
stepsperday <- summarize(activity_by_day, totalsteps = sum(steps))
print(stepsperday)
qplot(totalsteps, data=stepsperday, xlab="Total Step in One Day",
      ylab="Frequency", main="Histogram of Total Steps Per Day")
```

```{r}
meansteps <- mean(stepsperday$totalsteps, na.rm=T)
mediansteps <- median(stepsperday$totalsteps, na.rm=T)
```

**Mean Steps Per Day:** ```r format(meansteps, scientific=F)```  
**Median Steps Per Day:** ```r mediansteps```  

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
